Using Visual Analytics to Explain\\[1ex] Black-Box Machine Learning

### 2018-04-06 infuse/use_case.tex
typo fixing

\begin{figure*}[ht]
\centering
\resizebox{0.9\linewidth}{!}{%
\begin{tikzpicture}
\newcommand*{\imgwidth}{3.7cm}
\newcommand*{\boxsize}{0.437cm}
\newcommand*{\boxpos}{(.295, -.18)}
\newcommand*{\rightimagepos}{(4.8, 0)}
\newcommand*{\leftimagepos}{(0, 0)}
\node[inner sep=0pt] at \leftimagepos
    {\includegraphics[width=\imgwidth]{infuse/ap}};
\node[inner sep=0pt] at \leftimagepos
    {\phantom{\includegraphics[width=\imgwidth]{infuse/ap}}};
\node[inner sep=2.5pt,very thick, text height=\boxsize] (zoom)
at \boxpos
    {\hspace*{\boxsize}};
\node[inner sep=-0.05pt] (lasso) at \rightimagepos
    {\phantom{\includegraphics[width=\imgwidth]{infuse/lasso}}};
\draw (zoom.north west) -- (lasso.north west);
\draw (zoom.south west) -- (lasso.south west);
\draw (zoom.north east) -- (lasso.north east);
\draw (zoom.south east) -- (lasso.south east);
\node[very thick, draw=red!90, text height=\boxsize]
at \boxpos {\hspace*{\boxsize}};
\node[inner sep=0pt, very thick] at \rightimagepos
    {\includegraphics[width=\imgwidth]{infuse/lasso}};
\node[inner sep=0pt, very thick, draw=red!90] at \rightimagepos
    {\phantom{\includegraphics[width=\imgwidth]{infuse/lasso}}};
\end{tikzpicture}%
}
\caption[The scatter-plot view allows to compare multiple types of rankings.]{
The scatter-plot view allows users to compare multiple types of rankings.  In the case study, users became curious of the medication features that were chosen by only half of the models.  When reviewing these medications with domain experts, it became clear that features picked by the upper-half algorithms were as clinically significant as those picked by the bottom-half.  This indicates that merging results from feature selection algorithms makes sense for this predictive model.
}
\label{fig:usecase2}
\end{figure*}

\subsection{Insight 1: Data issues}
When the clinical researchers learned of \infuse's capabilities
to compare multiple feature selection algorithms, they decided to
expand their pipeline's feature selection algorithms from 2 to 4.
The team has used Information Gain and Fisher Score extensively
in prior work, and typically uses these same techniques due
to their familiarity and past success.
Nonetheless, the diabetes data set introduced in Section \ref{sec:running_example} was new to them, and they were unsure which techniques would be the most appropriate. So, they asked their technologists to implement two new techniques: Odds Ratio and Relative Risk.

@@

As \infuse gave them the opportunity to
examine multiple algorithms at the feature-level, they were curious as
to why this trend of non-overlapping feature rankings occurred.
They investigated the scores associated with each feature rank
and noticed that many of the features had scores
of $\infty$ from the Relative Risk algorithm.
It turned out there was a bug within the Relative Risk implementation where a
divide by zero error could happen if a feature did not occur
in any of the control patients.  After fixing this bug, they noticed that much of the non-overlap still was evident.  Looking more closely at the algorithms provided a
reason why the two new algorithms behave very differently:  they realized that both of the new algorithms only look at the presence and absence of the feature between the case and controls,
and do not pay attention to the feature values in any other way (e.g. distribution of values).
This is in contrast to the Fisher Score and Information Gain algorithms that take
the actual feature values into account.  This means that for features that are present in both case
and control groups in the same proportion, there is no discrimination value.

@@

As different algorithms will make sense for different purposes depending on the dataset and goals, \infuse provides an ability to inspect the features and determine
which algorithms produce ranked sets of domain-relevant features.

\subsection{Insight 2: Clinically relevant features}
After the data issues were solved, the researchers began investigating
the content of the predictive features.
Using the scatterplot view,
they inspected all of the medications that were ranked by all
feature selection algorithms and folds and found that they
were \textit{antihyperglycemic} medications, which are common treatments to lower the blood sugar of diabetes patients, and made clinical sense to be ranked high.

>>>>

\begin{figure}
\centering
\resizebox{0.9\linewidth}{!}{%
\begin{tikzpicture}
\newcommand*{\imgwidth}{3.7cm}
\newcommand*{\boxsize}{0.437cm}
\newcommand*{\boxpos}{(.295, -.18)}
\newcommand*{\rightimagepos}{(4.8, 0)}
\newcommand*{\leftimagepos}{(0, 0)}
\node[inner sep=0pt] at \leftimagepos
    {\includegraphics[width=\imgwidth]{infuse/ap}};
\node[inner sep=0pt] at \leftimagepos
    {\phantom{\includegraphics[width=\imgwidth]{infuse/ap}}};
\node[inner sep=2.5pt,very thick, text height=\boxsize] (zoom)
at \boxpos
    {\hspace*{\boxsize}};
\node[inner sep=-0.05pt] (lasso) at \rightimagepos
    {\phantom{\includegraphics[width=\imgwidth]{infuse/lasso}}};
\draw (zoom.north west) -- (lasso.north west);
\draw (zoom.south west) -- (lasso.south west);
\draw (zoom.north east) -- (lasso.north east);
\draw (zoom.south east) -- (lasso.south east);
\node[very thick, draw=red!90, text height=\boxsize]
at \boxpos {\hspace*{\boxsize}};
\node[inner sep=0pt, very thick] at \rightimagepos
    {\includegraphics[width=\imgwidth]{infuse/lasso}};
\node[inner sep=0pt, very thick, draw=red!90] at \rightimagepos
    {\phantom{\includegraphics[width=\imgwidth]{infuse/lasso}}};
\end{tikzpicture}%
}
\caption[The scatter-plot view allows to compare multiple types of rankings.]{
The scatter-plot view allows users to compare multiple types of rankings.  In the case study, users became curious of the medication features that were chosen by only half of the models.  When reviewing these medications with domain experts, it became clear that features picked by the upper-half algorithms were as clinically significant as those picked by the bottom-half.  This indicates that merging results from feature selection algorithms makes sense for this predictive model.
}
\label{fig:usecase2}
\end{figure}

\subsection{Insight 1: Data Issues}
When the clinical researchers learned of \infuse's capabilities
to compare multiple feature selection algorithms, they decided to
expand their pipeline's feature selection algorithms from 2 to 4.
The team has used Information Gain and Fisher Score extensively
in prior work, and typically uses these same techniques due
to their familiarity and past success.
Nonetheless, the diabetes data set introduced in Section \ref{sec:running_example} was new to them, and they were unsure which techniques would be the most appropriate. So, they asked their technologists to implement two new techniques: Odds Ratio and Relative Risk.

@@

As \infuse gave them the opportunity to
examine multiple algorithms at the feature-level, they were curious as
to why this trend of non-overlapping feature rankings occurred.
They investigated the scores associated with each feature rank
and noticed that many of the features had scores
of $\infty$ from the Relative Risk algorithm.
It turned out there was a bug within the Relative Risk implementation where a
divide by zero error could happen if a feature did not occur
in any of the control patients.  After fixing this bug, they noticed that much of the non-overlap still was evident.  Looking more closely at the algorithms provided a
reason why the two new algorithms behave very differently:  they realized that both of the new algorithms only look at the presence and absence of the feature between the case and controls,
and do not pay attention to the feature values in any other way (\eg, distribution of values).
This is in contrast to the Fisher Score and Information Gain algorithms that take
the actual feature values into account.  This means that for features that are present in both case
and control groups in the same proportion, there is no discrimination value.

@@

As different algorithms will make sense for different purposes depending on the data set and goals, \infuse provides an ability to inspect the features and determine
which algorithms produce ranked sets of domain-relevant features.

\subsection{Insight 2: Clinically Relevant Features}
After the data issues were solved, the researchers began investigating
the content of the predictive features.
Using the scatterplot view,
they inspected all of the medications that were ranked by all
feature selection algorithms and folds and found that they
were \textit{antihyperglycemic} medications, which are common treatments to lower the blood sugar of diabetes patients, and made clinical sense to be ranked high.

### 2018-04-06 prospector/case_studies.tex
typo fixing

In order to evaluate the utility of \prospector, we chose to conduct a case study utilizing a team of real data scientists building their own predictive models on their own real-world datasets to demonstrate its effectiveness at reaching insights in practice. There is a growing belief in the visualization community that traditional evaluation metrics (e.g. measuring task time completion or number of errors) are often insufficient to evaluate visualization systems \cite{bertini_summaries:_2011,plaisant_challenge_2004,shneiderman_strategies_2006}. Using the evaluation methodology developed by Perer and Shneiderman \cite{perer_integrating_2008}, we conducted a 4-month long-term case study with a team of five data scientists interested in using predictive modeling on a longitudinal database of electronic medical records. The research team is interested in building a predictive model to predict if a patient is at risk of developing diabetes using a database of 4,000 patients from a major hospital in the United States.  Due to sensitive data agreements, this team wished to remain anonymous.

@@

While all three models have a decrease in average predicted risk when the count goes from 0 to 2, the logistic regression models continue to trend downward.  However, the random forest model (in red), illustrates the predictive risk increases as the count gets higher than two.  The data scientists were surprised to learn how differently the model classes treated this feature, but using the tool, they were able to devise a two-fold explanation.  On one hand, logistic regression models are not expressive enough to model the late increase after the initial drop, as logistic regression models are bound by a single curve.  On the other hand, most of the observed data points of the feature are zero, one, and two while the higher values occur extremely rare, as the histogram below the plot clearly illustrates.  This led the data scientist team to question whether to model everything as precisely as possible or using a simpler model for the sake of generality.

@@

The last question about identifying actionable features was of particular importance to the data science team.  They were interested to know if the model could be used to learn features that could be acted upon by the patients or their doctors to reduce the risk of diabetes.  However, the data science team were disappointed to learn, via \prospector, that many of the highest ranking features were not actionable.  For instance, some of the most predictive features for a high risk of diabetes involved having a high count of the number of lab tests.  Informing patients to get fewer lab tests would likely not correlate to lower risk of diabetes.  Instead, these lab test counts were likely a proxy for other features that correlated to more complicated or more sickly patients seeing their doctors more regularly and thus getting more lab tests.  Other demographic features that were highly predictive, like age, simply have no intervention as well.  The data science team then reconsidered which features should be a part of the predictive model, by creating features that are actionable and omitting others.  Of course, the model cannot know this by itself -- no matter what sophisticated feature selection algorithms are used -- so the access \prospector provides is critical for this process.

\subsection{Impact into Data Scientists' Workflow}

@@

Once \prospector was integrated into their workflow, many of these shortcomings were overcome.  No longer did a data scientist need to generate a set of manual reports.  Instead, the predictive model can be loaded into the tool directly.  Since the tool is interactive, it also allows the team to ask questions that may have not been considered when static charts were created.  The tool also allowed them to ask questions beyond the top features that contributed to the models.  They could ask more patient-centric questions such as ``Why is this patient not being classified correctly?" by drilling down to incorrectly predicted patients and exploring the most impactful features for them. Beyond exploration, \prospector was also used to communicate models to the stakeholders, which allowed stakeholders to ask questions and see the results in real-time.  This rapid feedback helped gain support for deploying predictive models in future projects.  As a result of these successes, \prospector is now a part of their predictive modeling workflow and is used for other work than predicting the onset of diabetes.

>>>>

In order to evaluate the utility of \prospector, we chose to conduct a case study utilizing a team of real data scientists building their own predictive models on their own real-world datasets to demonstrate its effectiveness at reaching insights in practice. There is a growing belief in the visualization community that traditional evaluation metrics (\eg, measuring task time completion or number of errors) are often insufficient to evaluate visualization systems \cite{bertini_summaries:_2011,plaisant_challenge_2004,shneiderman_strategies_2006}. Using the evaluation methodology developed by Perer and Shneiderman \cite{perer_integrating_2008}, we conducted a 4-month long-term case study with a team of five data scientists interested in using predictive modeling on a longitudinal database of electronic medical records. The research team is interested in building a predictive model to predict if a patient is at risk of developing Diabetes using a database of 4,000 patients from a major hospital in the United States.  Due to sensitive data agreements, this team wished to remain anonymous.

@@

While all three models have a decrease in average predicted risk when the count goes from 0 to 2, the logistic regression models continue to trend downward.  However, the random forest model (in red), illustrates the predictive risk increases as the count gets higher than two.  The data scientists were surprised to learn how differently the model classes treated this feature, but using the tool, they were able to devise a two-fold explanation.  On one hand, logistic regression models are not expressive enough to model the late increase after the initial drop, as logistic regression models are bound by a single curve.  On the other hand, most of the observed instances of the feature are zero, one, and two while the higher values occur extremely rare, as the histogram below the plot clearly illustrates.  This led the data scientist team to question whether to model everything as precisely as possible or using a simpler model for the sake of generality.

@@

The last question about identifying actionable features was of particular importance to the data science team.  They were interested to know if the model could be used to learn features that could be acted upon by the patients or their doctors to reduce the risk of Diabetes.  However, the data science team were disappointed to learn, via \prospector, that many of the highest ranking features were not actionable.  For instance, some of the most predictive features for a high risk of Diabetes involved having a high count of the number of lab tests.  Informing patients to get fewer lab tests would likely not correlate to lower risk of Diabetes.  Instead, these lab test counts were likely a proxy for other features that correlated to more complicated or more sickly patients seeing their doctors more regularly and thus getting more lab tests.  Other demographic features that were highly predictive, like age, simply have no intervention as well.  The data science team then reconsidered which features should be a part of the predictive model, by creating features that are actionable and omitting others.  Of course, the model cannot know this by itself -- no matter what sophisticated feature selection algorithms are used -- so the access \prospector provides is critical for this process.

\subsection{Impact on Data Scientists' Workflow}

@@

Once \prospector was integrated into their workflow, many of these shortcomings were overcome.  No longer did a data scientist need to generate a set of manual reports.  Instead, the predictive model can be loaded into the tool directly.  Since the tool is interactive, it also allows the team to ask questions that may have not been considered when static charts were created.  The tool also allowed them to ask questions beyond the top features that contributed to the models.  They could ask more patient-centric questions such as ``Why is this patient not being classified correctly?" by drilling down to incorrectly predicted patients and exploring the most impactful features for them. Beyond exploration, \prospector was also used to communicate models to the stakeholders, which allowed stakeholders to ask questions and see the results in real-time.  This rapid feedback helped gain support for deploying predictive models in future projects.  As a result of these successes, \prospector is now a part of their predictive modeling workflow and is used for other work than predicting the onset of Diabetes.

### 2018-04-06 prospector/conclusion.tex
typo fixing

In this paper, we demonstrated how the design and implementation of an interactive visual analytics system,
\prospector, can help data scientists assess the interpretability and actionable insights of trained predictive models. \prospector accomplishes this by supporting interactive partial dependence diagnostics for understanding how features affect the prediction overall by featuring novel visual representations, sampling strategies, and support for comparing multiple models.  Furthermore, \prospector supports localized inspection so data scientists can understand how and why specific datapoints are predicted as they are.  With support to interactively tweak feature values and see how the prediction responds, as well as finding the most impactful features using a novel local feature importance metric, data scientists can interact with models on a deeper level than possible with common tools.  Finally, we presented a case study, in the spirit of \#chi4good, which involved a team of data scientists using \prospector  to improve predictive models for detecting the onset of diabetes.  Their extended use of the tool led to better predictive models, as well as better communication of their models to their stakeholders.

@@

Despite the novel features of \prospector and its successful case study, there is still much future work to continue to give users full comprehension of predictive models. \prospector relies on partial dependence for one input feature at a time, but this approach relies on the orthogonality of input features.  However, in real-world data, this is not often the case, as features may be correlated.  \prospector can only model changes along one axis at a time as it cannot take correlations or influences between features into account.
We plan to address this limitation in future work by modeling valid sets of input points and visualizing how
they react to changes in one or more features.  Another limitation is that \prospector was built to view predictive models after they had been built using users' own predictive modeling pipeline of choice.  However, this flexibility limits the ability for users to directly impact their predictive models based on insights reached during exploration. Also, \prospector can only handle single-class predictions, but we plan to extend this functionality to multi-class predictions in the future.
Our future work also intends to integrate \prospector more directly into the predictive modeling pipeline so users can directly modify features for feature construction and feature selection and see how their models improve in a single user interface.  Despite these limitations, providing users with advanced visual tools to inspect black-boxes of machine learning shows great promise and helps users comprehend and retain control of their predictive models without sacrificing accuracy.

>>>>

In this paper, we demonstrated how the design and implementation of an interactive visual analytics system,
\prospector, can help data scientists assess the interpretability and actionable insights of trained predictive models. \prospector accomplishes this by supporting interactive partial dependence diagnostics for understanding how features affect the prediction overall by featuring novel visual representations, sampling strategies, and support for comparing multiple models.  Furthermore, \prospector supports localized inspection so data scientists can understand how and why specific instances are predicted as they are.  With support to interactively tweak feature values and see how the prediction responds, as well as finding the most impactful features using a novel local feature importance metric, data scientists can interact with models on a deeper level than possible with common tools.  Finally, we presented a case study, in the spirit of \#chi4good, which involved a team of data scientists using \prospector  to improve predictive models for detecting the onset of Diabetes.  Their extended use of the tool led to better predictive models, as well as better communication of their models to their stakeholders.

@@

Despite the novel features of \prospector and its successful case study, there is still much future work to continue to give users full comprehension of predictive models. \prospector relies on partial dependence for one input feature at a time, but this approach relies on the orthogonality of input features.  However, in real-world data, this is not often the case, as features may be correlated.  \prospector can only model changes along one axis at a time as it cannot take correlations or influences between features into account.
We plan to address this limitation in future work by modeling valid sets of instances and visualizing how
they react to changes in one or more features.  Another limitation is that \prospector was built to view predictive models after they had been built using users' own predictive modeling pipeline of choice.  However, this flexibility limits the ability for users to directly impact their predictive models based on insights reached during exploration. Also, \prospector can only handle single-class predictions, but we plan to extend this functionality to multi-class predictions in the future.
Our future work also intends to integrate \prospector more directly into the predictive modeling pipeline so users can directly modify features for feature construction and feature selection and see how their models improve in a single user interface.  Despite these limitations, providing users with advanced visual tools to inspect black-boxes of machine learning shows great promise and helps users comprehend and retain control of their predictive models without sacrificing accuracy.

### 2018-04-06 prospector/introduction.tex
typo fixing

In the era of data-driven analytics, there is growing demand to generate and deploy predictive models in a variety of domains so that the patterns unearthed from massive amounts of data can be leveraged
% for the greater good.
and converted into actionable insights.
Predictive modeling is defined as the process of developing a mathematical tool or model that generates an accurate prediction \cite{kuhn2013applied}.  As an example, in healthcare, if one can model the data characteristics of patients who will likely develop diabetes, healthcare institutions could deploy such a model on their patient databases, and automatically flag high risk patients to clinicians to make sure they are being treated appropriately.  However, building such models on noisy, real-world data is quite challenging.

@@

\begin{itemize}
\item A design and implementation of an interactive visual analytics system,
\prospector, for assessing the interpretability and actionable insights of trained predictive models by supporting:
\begin{itemize}
    \item Interactive partial dependence diagnostics to understand how features affect the prediction overall. There are novel visual representations, sampling strategies, and support for comparing multiple models.
    \item Localized inspection to understand how and why specific data points are predicted as they are.  Users can interactively tweak feature values and see how the prediction responds, as well as find the most impactful features using a novel local feature importance metric.
  \end{itemize}
\item A case study of data scientists using
\prospector to improve predictive models for detecting the onset of diabetes trained from electronic medical record data.
\end{itemize}

>>>>

In the era of data-driven analytics, there is growing demand to generate and deploy predictive models in a variety of domains so that the patterns unearthed from massive amounts of data can be leveraged
% for the greater good.
and converted into actionable insights.
Predictive modeling is defined as the process of developing a mathematical tool or model that generates an accurate prediction \cite{kuhn2013applied}.  As an example, in health care, if one can model the data characteristics of patients who will likely develop Diabetes, health care institutions could deploy such a model on their patient databases, and automatically flag high risk patients to clinicians to make sure they are being treated appropriately.  However, building such models on noisy, real-world data is quite challenging.

@@

\begin{itemize}
\item A design and implementation of an interactive visual analytics system,
\prospector, for assessing the interpretability and actionable insights of trained predictive models by supporting:
\begin{itemize}
    \item Interactive partial dependence diagnostics to understand how features affect the prediction overall. There are novel visual representations, sampling strategies, and support for comparing multiple models.
    \item Localized inspection to understand how and why specific instances are predicted as they are.  Users can interactively tweak feature values and see how the prediction responds, as well as find the most impactful features using a novel local feature importance metric.
  \end{itemize}
\item A case study of data scientists using
\prospector to improve predictive models for detecting the onset of Diabetes trained from electronic medical record data.
\end{itemize}

### 2018-04-06 prospector/local_inspection.tex
typo fixing

Our second core contribution is leveraging our implementation of partial dependence to support the user task of local inspection.  Users can use \prospector to inspect specific data points of interest and see how the models predict how they behave.  In addition, if the users are curious about how a particular data point's risk might change if it had different values, a user can explore this as well.  The idea of localized inspection is illustrated in Figure~\ref{figs:liexplain} using our running example of diabetes prediction.  At the top, the original patient's feature values are shown, along with the patient's original predicted low risk of having diabetes. Suppose the analyst was curious to see how the patient's risk would change if his BMI was increased to 35.
Localized inspection allows users to interactively change this value, and see the corresponding prediction.
In order to streamline this kind of exploration we fully compute the
predicted risk for all values of BMI similar to partial dependence.
As seen in Figure~\ref{figs:liexplain} we do this for all features independently
yielding local partial dependence plots for each feature using a single input row.
% Using the idea of partial dependence on a single row of the input data set creates a localized partial dependence describing the axis aligned neighborhood of one point in the prediction function. This enables what-if scenarios by changing some values of the input row and seeing how the predicted outcome changes.

@@

\subsubsection{Local feature importance}
The fourth novel contribution of our research is a technique to simplify the
exploration of the predicted risk space by automatically finding features where a small change in value yields a significantly large change on the predicted risk.

>>>>

Our second core contribution is leveraging our implementation of partial dependence to support the user task of local inspection.  Users can use \prospector to inspect specific data points of interest and see how the models predict how they behave.  In addition, if the users are curious about how a particular data point's risk might change if it had different values, a user can explore this as well.  The idea of localized inspection is illustrated in Figure~\ref{figs:liexplain} using our running example of Diabetes prediction.  At the top, the original patient's feature values are shown, along with the patient's original predicted low risk of having Diabetes. Suppose the analyst was curious to see how the patient's risk would change if his BMI was increased to 35.
Localized inspection allows users to interactively change this value, and see the corresponding prediction.
In order to streamline this kind of exploration we fully compute the
predicted risk for all values of BMI similar to partial dependence.
As seen in Figure~\ref{figs:liexplain} we do this for all features independently
yielding local partial dependence plots for each feature using a single input row.
% Using the idea of partial dependence on a single row of the input data set creates a localized partial dependence describing the axis aligned neighborhood of one point in the prediction function. This enables what-if scenarios by changing some values of the input row and seeing how the predicted outcome changes.

@@

\subsubsection{Local Feature Importance}
The fourth novel contribution of our research is a technique to simplify the
exploration of the predicted risk space by automatically finding features where a small change in value yields a significantly large change on the predicted risk.

### 2018-04-06 prospector/ml_crash.tex
typo fixing

A decision tree is one such algorithm, which is a tree whose nodes are rules that decide how to proceed down the branches of the tree, according to the range of values for a specific feature.  The decision making starts at the root of the tree and leaves carry the prediction results. Decision trees are popular in machine learning as they allow data scientists to model arbitrary functions.
However, the more nodes a tree has, the harder it is to understand the reasoning behind outcomes.
Logistic regression is another popular algorithm that is easier to understand, as features can only positively or negative influence the prediction, and the rate of influence is fixed.  This is achieved by defining a hyper-plane in the feature vector space, where the outcome of the prediction depends on how close to and on which side of the hyper-plane a point is.

>>>>

A decision tree is one such algorithm, which is a tree whose nodes are rules that decide how to proceed down the branches of the tree, according to the range of values for a specific feature.  The decision making starts at the root of the tree and leaves carry the prediction results. Decision trees are popular in machine learning as they allow data scientists to model arbitrary functions.
However, the more nodes a tree has, the harder it is to understand the reasoning behind outcomes.
Logistic regression is another popular algorithm that is easier to understand, as features can only positively or negative influence the prediction, and the rate of influence is fixed.  This is achieved by defining a hyper-plane in the feature vector space, where the outcome of the prediction depends on how close to and on which side of the hyper-plane an instance is.

### 2018-04-06 prospector/motivation.tex
typo fixing

\subsection{Predictive Modeling in Healthcare}

In order to make our contributions concrete, we utilize a motivating scenario that emerged from our case study.  The case study involves a team of five data scientists interested in using predictive modeling on a longitudinal database of electronic medical records. The research team has a background in healthcare analytics and their database contains 4,000 patients from a major hospital in the United States. The team is interested in building a predictive model to predict if a patient is at risk of developing diabetes, a chronic disease of high blood sugar levels that may cause serious health complications.

@@

The team of data scientists manages to develop a highly accurate predictive model for detecting patients at high risk of developing diabetes.  They determine its effectiveness by measuring the common metrics used by predictive models (e.g., accuracy and AUC scores \cite{kuhn2013applied}).  They also followed the best practices of building predictive models.  They worked with clinical researchers to properly define cohorts of patients with diabetes (cases) and matched patients without diabetes (controls) by thoroughly searching through the electronic medical records.  They constructed features based on lab tests, diagnosis codes, procedures, demographics, and patient conditions from the records.  They used cross-validation to ensure their models were robust.  They used a variety of state-of-the-art feature selection methods to utilize the most informative features in the model while keeping it as generalizable as possible.  And they used a variety of effective classifiers to do the training and evaluation.  After trying various combinations of these techniques, the model with the highest accuracy metrics was selected and presented to the appropriate stakeholders at the healthcare institution.

Their stakeholders were impressed by the high accuracy scores of the model.  But when they asked the data scientists for more information about what was inside of the model, the reports only described the top features of the model and their associated ``importance" weights.  The stakeholders recognized many of the feature names, and it appeared to make clinical sense.  However, there were also some surprising ones that led to intellectual discussions.  But the stakeholders demanded to know more.  They wanted a clearer sense of how certain features impacted the prediction.  Furthermore, they wanted to understand how the values associated with the features (e.g., the results associated with lab tests) impacted the prediction.  They also were curious to interact with the model to test hypotheses about how the model would react to hypothetical patients of interest.  When confronted with these questions, the data scientists shrugged.  In the data scientist's defense, it is difficult to summarize and interpret complex models and there are few tools and techniques available to address the stakeholders' requests.  So now the stakeholders are faced with a hard decision:  do they deploy a predictive model in their institution that appears to have high accuracy but is also somewhat of a black-box?

>>>>

\subsection{Predictive Modeling in Health Care}

In order to make our contributions concrete, we utilize a motivating scenario that emerged from our case study.  The case study involves a team of five data scientists interested in using predictive modeling on a longitudinal database of electronic medical records. The research team has a background in health care analytics and their database contains 4,000 patients from a major hospital in the United States. The team is interested in building a predictive model to predict if a patient is at risk of developing Diabetes, a chronic disease of high blood sugar levels that may cause serious health complications.

@@

The team of data scientists manages to develop a highly accurate predictive model for detecting patients at high risk of developing Diabetes.  They determine its effectiveness by measuring the common metrics used by predictive models (e.g., accuracy and AUC scores \cite{kuhn2013applied}).  They also followed the best practices of building predictive models.  They worked with clinical researchers to properly define cohorts of patients with Diabetes (cases) and matched patients without Diabetes (controls) by thoroughly searching through the electronic medical records.  They constructed features based on lab tests, diagnosis codes, procedures, demographics, and patient conditions from the records.  They used cross-validation to ensure their models were robust.  They used a variety of state-of-the-art feature selection methods to utilize the most informative features in the model while keeping it as generalizable as possible.  And they used a variety of effective classifiers to do the training and evaluation.  After trying various combinations of these techniques, the model with the highest accuracy metrics was selected and presented to the appropriate stakeholders at the health care institution.

Their stakeholders were impressed by the high accuracy scores of the model.  But when they asked the data scientists for more information about what was inside of the model, the reports only described the top features of the model and their associated ``importance" weights.  The stakeholders recognized many of the feature names, and it appeared to make clinical sense.  However, there were also some surprising ones that led to intellectual discussions.  But the stakeholders demanded to know more.  They wanted a clearer sense of how certain features impacted the prediction.  Furthermore, they wanted to understand how the values associated with the features (\eg, the results associated with lab tests) impacted the prediction.  They also were curious to interact with the model to test hypotheses about how the model would react to hypothetical patients of interest.  When confronted with these questions, the data scientists shrugged.  In the data scientist's defense, it is difficult to summarize and interpret complex models and there are few tools and techniques available to address the stakeholders' requests.  So now the stakeholders are faced with a hard decision:  do they deploy a predictive model in their institution that appears to have high accuracy but is also somewhat of a black-box?

### 2018-04-06 prospector/partial_dependence.tex
typo fixing

In order to make this function more concrete, consider the illustrative example in Figure~\ref{figs:pdexplain}, where each input row is a patient, and each column is a feature.  The last column represents the output of the predictive model that predicts if a patient is at low-risk or high-risk for developing diabetes.  In Figure~\ref{figs:pdexplain}a, the patients' original feature values (age, BMI (Body Mass Index, a standard way to quantify obesity), and glucose level (a standard way to determine diabetes)) are shown.  If one wants to examine the impact of the glucose feature on the prediction, partial dependence can be applied by keeping all of the other features (age, bmi) as they were, but fixing glucose to a set of fixed values to see how that feature impacts the prediction.  For example, in Figure~\ref{figs:pdexplain}b, the glucose values (highlighted in yellow) are fixed to 100, which yields predictions of only 1 patient being high risk, instead of the original 2.  Conversely, in Figure~\ref{figs:pdexplain}c, glucose is fixed to 180, and 3 patients are predicted to have high risk.  Thus, there appears to be partial dependence of glucose on the prediction.

Partial dependence is typically visualized as a partial dependence plot, as shown in Figure~\ref{figs:pdp},which is a line graph that plots the fixed values for the target feature on the x-axis, and the corresponding predicted risk (probability of a certain outcome) on the y-axis.

>>>>

In order to make this function more concrete, consider the illustrative example in Figure~\ref{figs:pdexplain}, where each input row is a patient, and each column is a feature.  The last column represents the output of the predictive model that predicts if a patient is at low-risk or high-risk for developing Diabetes.  In Figure~\ref{figs:pdexplain}a, the patients' original feature values (age, BMI (Body Mass Index, a standard way to quantify obesity), and glucose level (a standard way to determine Diabetes)) are shown.  If one wants to examine the impact of the glucose feature on the prediction, partial dependence can be applied by keeping all of the other features (age, BMI) as they were, but fixing glucose to a set of fixed values to see how that feature impacts the prediction.  For example, in Figure~\ref{figs:pdexplain}b, the glucose values (highlighted in yellow) are fixed to 100, which yields predictions of only 1 patient being high risk, instead of the original 2.  Conversely, in Figure~\ref{figs:pdexplain}c, glucose is fixed to 180, and 3 patients are predicted to have high risk.  Thus, there appears to be partial dependence of glucose on the prediction.

Partial dependence is typically visualized as a partial dependence plot, as shown in Figure~\ref{figs:pdp}, which is a line graph that plots the fixed values of the target feature on the x-axis, and the corresponding predicted risk (\ie, probability of a certain outcome) on the y-axis.
